Overlooked Dynamics in Algorithmic Recourse

### 2022-07-07 paper/sections/methodology.qmd
minor changes to paper draft

The dynamics illustrated in @fig-poc in @sec-intro were generated through a simple experiment that aims to simulate the process of algorithmic recourse in practice. We begin in the static setting at time time $t=0$: given some pre-trained classifier $M$ we generate recourse for a random batch of $B$ individuals in the non-target class. Note that we focus our attention on classification problems, since classification poses the most common practical use-case for algorithmic recourse. In order to simulate the dynamical process we suppose that the model $M$ is retrained following the actual implementation of recourse in time $t=0$. Following the update to the model, we assume that at time $t=1$ recourse is generated for yet another random subset of individuals in the non-target class. This process is repeated for a number of time periods $T$. To get a clean read on endogenous dynamics we keep the total population of samples closed: we allow existing samples to move from factual to counterfactual states, but do not allow any entirely new samples to enter the population. The experimental setup is summarized in Algorithm \ref{euclid}

\begin{algorithm}
\caption{PLACEHOLDER}\label{euclid}
\begin{algorithmic}[1]
\Procedure{Euclid}{$a,b$}\Comment{The g.c.d. of a and b}
\State $r\gets a\bmod b$
\While{$r\not=0$}\Comment{We have the answer if r is 0}
\State $a\gets b$
\State $b\gets r$
\State $r\gets a\bmod b$
\EndWhile\label{euclidendwhile}
\State \textbf{return} $b$\Comment{The gcd is b}
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

![A visualization of the synthetic classification datasets used in our experiments.](www/synthetic_data.png){#fig-synthetic-data}

@@

>>>>

The dynamics illustrated in @fig-poc in @sec-intro were generated through a simple experiment that aims to simulate the process of algorithmic recourse in practice. We begin in the static setting at time time $t=0$: given some pre-trained classifier $M$ we generate recourse for a random batch of $B$ individuals in the non-target class. Note that we focus our attention on classification problems, since classification poses the most common practical use-case for algorithmic recourse. In order to simulate the dynamical process we suppose that the model $M$ is retrained following the actual implementation of recourse in time $t=0$. Following the update to the model, we assume that at time $t=1$ recourse is generated for yet another random subset of individuals in the non-target class. This process is repeated for a number of time periods $T$. To get a clean read on endogenous dynamics we keep the total population of samples closed: we allow existing samples to move from factual to counterfactual states, but do not allow any entirely new samples to enter the population. The experimental setup is summarized in Algorithm \ref{algo-experiment}

\begin{algorithm}
\caption{Experiment}\label{algo-experiment}
\begin{algorithmic}[1]
\Procedure{Experiment}{$M,D,G$}
\State $t\gets 0$
\While{$t<T$}
\State $D_B \subset D$
\State $D_B\gets G(D_B)$ \Comment{Generate counterfactuals.}
\State $M\gets M(D)$ \Comment{Retrain model.}
\EndWhile
\State \textbf{return} $M,D$
\EndProcedure
\end{algorithmic}
\end{algorithm}

@@

![PLACEHOLDER: A visualization of the synthetic classification datasets used in our experiments.](www/synthetic_data.png){#fig-synthetic-data}

@@

As our baseline for quantifying model shifts we measure perturbations to the model parameters at each point in time $t$ following @upadhyay2021towards. We define $\Delta=||\theta_{t+1}-\theta_{t}||^2$, that is the euclidean distance between the vectors of parameters before and after retraining the model $M$. We shall refer to this baseline metric simply as **Perturbations**.
